---
title: "Homework"
author: "Arielle Glass"
output: html_document
date: "2024-02-12"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#### Load libraries

```{r}
library(tidyverse)
library(dplyr)
library(tidyr)
```

#### Read in data 

```{r}
#Load data
surveys <- read_csv("diversity.csv")
```

### Tidyverse Day 1

#### Q1: What is the class of the object surveys? 

```{r}
#Q1
class(surveys)
str(surveys)
```

Surveys is a dataframe.

#### Q2: How many rows and how many columns are in the object?

Rows: 34,786
Columns: 13

#### Q2: Create a data.frame (surveys_200) containing only the data in row 200 of the surveys dataset.

```{r}
#Q2
surveys_200 <- surveys[200, ]
print(surveys_200)

```

#### Q3: Create a data.frame (surveys_last) containing only the data in the last row of the surveys dataset using nrow() instead of the row number. 

```{r}
#Q3
nrow <- nrow(surveys)
# number of rows/columns
surveys_last <- surveys[nrow, ] 
print(surveys_last)

```

#### Q4: Change the columns taxa and genus in the surveys data frame into a factor.

```{r}
#Q4
surveys$taxa <- factor(surveys$taxa)
surveys$genus <- factor(surveys$genus)
str(surveys)
```

#### Q5: Using the functions you learned before

##### Q5 A: How many rabbits were observed? 

```{r}
#Q5A
summary(surveys)

```
There are 75 rabbits.

##### Q5 B: How many different genera are in the genus column?

```{r}
#Q5B

length(unique(surveys$genus))
```

There are 26 genera.

#### Q6: Create new variable 'sex' by converting surveys$sex to a factor. Use addNA to add a factor level to 'sex' variable. Using the 'sex' variable, rename "NA", "F", and "M" factors to "undetermined", "female", and "male" respectively. Plot sex variable. 

```{r}
#Q6
sex <- factor(surveys$sex)
sex <- addNA(sex)
levels(sex)
levels(sex) <- c("undetermined", "female", "male")

plot(sex)

```

#### Q7: Recreate the barplot such that "undetermined" is first (before "female").

```{r}
#Q7
sex <- factor(sex, levels = c("undetermined", "female", "male"))
plot(sex)

```

#### Q8: Using pipes, subset the surveys data to include animals collected before 1995 and retain only the columns year, sex, and weight. View the tail() of the data.frame.

```{r}
#Q8
surveys_2 <- surveys %>%
  filter(year < 1995) %>%
  select(year, sex, weight)

tail(surveys_2)

```

### Tidyverse Day 2

#### Q9: Create a new data frame from the surveys data that meets the following criteria:

##### Contains only the species_id column and a new column called hindfoot_cm which contains the hindfoot_length values (currently in mm) converted to centimeters. In this hindfoot_cm column, there are no NAs and all values are less than 3.\

#### View the head() of the data.frame. 

##### Hint: Think about how the commands should be ordered to produce this data frame! 

```{r}
#Q9 

surveys_3 <- surveys %>%
  mutate(hindfoot_cm = hindfoot_length / 10) %>%
  filter(!is.na(hindfoot_cm) & hindfoot_cm < 3) %>%
  select(species_id, hindfoot_cm)

head(surveys_3)
 
```

#### Q10: How many animals were caught in each plot_type surveyed?

```{r}
#Q10

animals_per_plot_type <- surveys %>%
  group_by(plot_type) %>%
  summarize(count = n())

print(animals_per_plot_type)
```

#### Q11: Use group_by() and summarize() to find the mean, min, and max hindfoot length for each species (using species_id). Also add the number of observations (hint: see ?n).

```{r}
#Q11

hindfoot_stats_per_species <- surveys %>%
  group_by(species_id) %>%
  summarize(
    mean_hindfoot_length = mean(hindfoot_length, na.rm = TRUE),
    min_hindfoot_length = min(hindfoot_length, na.rm = TRUE),
    max_hindfoot_length = max(hindfoot_length, na.rm = TRUE),
    number_of_observations = n()
  )

print(hindfoot_stats_per_species)
```

#### Q12: What was the heaviest animal measured in each year? Return the columns year, genus, species_id, and weight.

```{r}
#Q12

heaviest_animal_per_year <- surveys %>%
  # Filter rows with non-missing weight
  filter(!is.na(weight)) %>%
  group_by(year) %>%
  summarize(
    genus = genus[which.max(weight)],
    species_id = species_id[which.max(weight)],
    weight = max(weight)
  )

print(heaviest_animal_per_year)
```

#### Q13: Reshape the surveys data frame with year as columns, plot_id as rows, and the number of genera per plot as the values. You will need to summarize before reshaping, and use the function n_distinct() to get the number of unique genera within a particular chunk of data. It’s a powerful function! See ?n_distinct for more.

```{r}
#Q13

reshaped_data <- surveys %>%
  group_by(plot_id, year) %>%
  summarize(num_genera = n_distinct(genus)) %>%
  pivot_wider(names_from = year, values_from = num_genera)

print(reshaped_data)
```

#### Q14: Now take that data frame and pivot_longer() it, so each row is a unique plot_id by year combination.

```{r}
#Q14

longer_data <- reshaped_data %>%
  pivot_longer(
    cols = -plot_id, 
    names_to = "year",
    values_to = "num_genera",
    names_transform = list(year = as.integer)
  )

print(longer_data)

```





